Revealing the rich and varied dynamics of hippocampal replay

### 2020-09-02 Article.tex

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[6]{Kenneth Kay}
\author[5]{Marielena Sosa}
\author[7]{Jason E. Chung}
\author[4]{Uri T. Eden}
\author[1, 2, 3, \Letter]{Loren M. Frank}

\affil[1]{Howard Hughes Medical Institute, University of California, San Francisco, San Francisco, California}
\affil[2]{Department of Physiology, University of California, San Francisco, San Francisco, California}
\affil[3]{Kavli Institute for Fundamental Neuroscience, University of California, San Francisco, San Francisco, California}
\affil[4]{Department of Mathematics and Statistics, Boston University, Boston, Massachusetts}
\affil[5]{Department of Neurobiology, Stanford University School of Medicine, Stanford, CA}
\affil[6]{Mortimer B. Zuckerman Mind Brain Behavior Institute, Columbia University, New York, New York}
\affil[7]{Department of Neurological Surgery, University of California, San Francisco, California}

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For the second 60 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. For the last 60 ms, cells fire randomly, out of their spatial order, representing an incoherent spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the posterior from the previous time step. In the stationary movement dynamic, the transition matrix is modeled as the identity matrix because the position remains the same. In the continuous movement dynamic, the transition matrix is modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close to the previous position. In the fragmented movement dynamic, the transition matrix is modeled as uniform, meaning the next position is equally likely to be any possible position. Note that the movement dynamics displayed only reflect the within-state movement dynamics. \textbf{(D)} The probability of remaining in a particular dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time. The summed posterior across states also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. The lines correspond to the average probability of that dynamic for a given constant velocity sequence. We use a 0.8  threshold (dotted line) to classify each sequence based on the state or states which contribute maximally to the posterior. 
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 180 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 60 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 60 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

Next, we specify a movement dynamics model for how latent position---the "mental" position of the animal represented by the cells---evolves over time (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can move from the previous time step. Previous findings suggest that replay may exhibit at least three distinct types of movement: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, rapid continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another. We therefore define state transition models to capture each of these dynamics.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v4}
\caption{A-D. Four examples of replays that would not be well-characterized by using the standard decoder. Figure conventions are the same as in Figure \ref{2}. \textbf{(A)} A replay that starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that persistently represents the left well for the entirety of the SWR while the animal remains in the center arm. \textbf{(C)} A replay that begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each epoch that has some content that falls into one of the categories of dynamics (Stationary, Stationary-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each epoch that have some Stationary, Stationary-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each epoch that has some Continuous content.
}
\label{3}
\end{figure*}

@@

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4511 of 23382 or 19\% of classified SWRs had some continuously moving content (Figure \ref{3}G, 17\% of all SWRs). Thus, focusing on only spatially continuous trajectories misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

@@

Our model is specified in the context of a latent position associated with different movement dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) region of the posterior linear position. Larger values of the cumulative HPD extent indicate the model is less certain about position, because the HPD values are distributed over more of the track at a given time point. In contrast, lower values indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD provides a complementary measure of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

We found that spatially coherent events did indeed have lower cumulative HPD sizes than events with fragmented dynamics. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD size is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD size is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the cumulative HPD was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent).

@@

Our results may also explain the seeming conflict between Stella and colleagues \cite{StellaHippocampalReactivationRandom2019}, who reported that hippocampal replay trajectories followed Brownian-like diffusion dynamics, and Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015}, who reported that replay trajectories contained jumps in position. We found that while replay speeds are on average distributed log normally, there are still definitive "jumps" from one position to another in a small subset of the replays. Our model allows identification of both types of sequences. Our model also makes it possible to identify different dynamics within the same SWR. As the start and end time of each SWR event are also dependent on a selection of arbitrary parameters, we suggest that instead, evaluating the start and end of each dynamic as an alternative landmark system may prove a powerful way identify periods of interest. Examination of such periods may shed light on when and why the hippocampus may be representing current position, past experience, or possible future locations, to understand how inputs to the hippocampus influence these events, and how these events may in turn influence downstream structures.

@@

Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated \cite{ChenBayesiannonparametricmethods2016, KloostermanBayesiandecodingusing2014, DengRapidclassificationhippocampal2016}. Third, we built an explicit, but flexible model that allowed us to identify not just one dynamic, but multiple dynamics consistent with different speeds of motion. Fourth, our model avoids assumptions about temporal bins and does not make assumptions about the linear progression of events. Instead, the model can use any size of bin and allows for movement in any direction. Finally, we used an acausal estimator of latent position, which allowed us to identify the specific timing of transitions in dynamics at a fine timescale.

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[4]{Kenneth Kay}
\author[5]{Marielena Sosa}
\author[6]{Jason E. Chung}
\author[7]{Uri T. Eden}
\author[1, 2, 3, \Letter]{Loren M. Frank}

\affil[1]{Howard Hughes Medical Institute, University of California, San Francisco, San Francisco, California}
\affil[2]{Departments of Physiology and Psychiatry, University of California, San Francisco, San Francisco, California}
\affil[3]{Kavli Institute for Fundamental Neuroscience, University of California, San Francisco, San Francisco, California}
\affil[4]{Mortimer B. Zuckerman Mind Brain Behavior Institute, Columbia University, New York, New York}
\affil[5]{Department of Neurobiology, Stanford University School of Medicine, Stanford, CA}
\affil[6]{Department of Neurological Surgery, University of California, San Francisco, California}
\affil[7]{Department of Mathematics and Statistics, Boston University, Boston, Massachusetts}

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Verifying the model can capture different sequence dynamics on simulated data and showing how it can classify different trajectory speeds. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For the second 60 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. For the last 60 ms, cells fire randomly, out of their spatial order, representing an incoherent spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the posterior from the previous time step. In the stationary movement dynamic, the transition matrix is modeled as the identity matrix because the position remains the same. In the continuous movement dynamic, the transition matrix is modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close to the previous position. In the fragmented movement dynamic, the transition matrix is modeled as uniform, meaning the next position is equally likely to be any possible position. Note that the movement dynamics displayed only reflect the within-state movement dynamics. \textbf{(D)} The probability of remaining in a particular dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time. The summed posterior across states also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. The lines correspond to the average probability of that dynamic for a given constant velocity sequence. We use a 0.8  threshold (dotted line) to classify each sequence based on the state or states which contribute maximally to the posterior. 
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 180 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 60 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 60 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

Next, we specify a movement dynamics model for how latent position---the "mental" position of the animal represented by the cells---evolves over time (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can move from the previous time step. Previous findings suggest that replay may exhibit at least three distinct types of movement: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, rapid continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \cite{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these dynamics.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v4}
\caption{A-D. Four examples of replays that would not be well-characterized by using the standard decoder. Figure conventions are the same as in Figure \ref{2}. \textbf{(A)} A replay that starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that persistently represents the left well for the entirety of the SWR while the animal remains in the center arm. \textbf{(C)} A replay that begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each day that has some content that falls into one of the categories of dynamics (Stationary, Stationary-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each day that have some Stationary, Stationary-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each day that has some Continuous content.
}
\label{3}
\end{figure*}

@@

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4511 of 23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G, 17\% of all SWRs). Thus, focusing on only high speed trajectories misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

@@

Our model is specified in the context of a latent position associated with different movement dynamics. This specification allows us to not only to classify events in terms of their dynamics, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) region of the posterior linear position. Larger values of the cumulative HPD extent indicate the model is less certain about position, because the HPD values are distributed over more of the track at a given time point. In contrast, lower values indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD provides a complementary measure of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates. Although this is certainly influenced by the dynamics imposed on the model via the movement state transition matrix, because the model is driven by the data, it does not necessarily have to be the case that the

We found that spatially coherent events did indeed have lower cumulative HPD sizes than events with fragmented dynamics. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD size is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD size is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E,). For the spatially coherent dynamics, the cumulative HPD was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U).

@@

Our results may also explain the seeming conflict between Stella and colleagues \cite{StellaHippocampalReactivationRandom2019}, who reported that hippocampal replay trajectories followed Brownian-like diffusion dynamics, and Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015}, who reported that replay trajectories contained jumps in position. We found that while replay speeds are on average distributed log normally, there are still definitive "jumps" from one position to another in a small subset of the replays. Our model allows identification of both types of sequences. Our model also makes it possible to identify different dynamics within the same SWR. As the start and end time of each SWR event are also parameter-dependent, we suggest that instead, evaluating the start and end of each dynamic as an alternative landmark system may prove a powerful way identify periods of interest. Examination of such periods may shed light on when and why the hippocampus may be representing current position, past experience, or possible future locations, to understand how inputs to the hippocampus influence these events, and how these events may in turn influence downstream structures.

@@

Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated \cite{ChenBayesiannonparametricmethods2016, KloostermanBayesiandecodingusing2014, DengRapidclassificationhippocampal2016}. Third, we built an explicit, but flexible model that allowed us to identify not just one dynamic, but multiple dynamics consistent with different speeds of motion. And we use these dynamics to describe the SWR events, rather than declaring the events as significant or not. Fourth, our model avoids assumptions about temporal bins and does not make assumptions about the linear progression of events. Instead, the model can use any size of bin and allows for movement in any direction. Finally, we used an acausal estimator of latent position, which allowed us to identify the specific timing of transitions in dynamics at a fine timescale. Our method allows us to identify periods of slower moving events that
